Title Goes Here \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2021-04-07 analysis/paper/shceof.tex
Renderiza todo

\begin{figure}
\centering
\includegraphics{../figures/eof-naive-1.pdf}
\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units). The numbers at the top-left of each panel ate the zonal variance explained by each EOF.}
\end{figure}

@@

This is analogous to the method of \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure~\ref{fig:phase-histogram} with their Figure 6).

\textcolor{red}{(Amplia un poco la oración anterior que no queda del todo clara)}

@@

\begin{figure}
\centering
\includegraphics{../figures/sam-eof-vertical-1.pdf}
\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}. Points mark estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
\end{figure}

To explore the relationship between SAM and the modes described from the cEOF, it is computed the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.~\ref{fig:sam-eof-vertical}). Both cEOFs bear some modest relationship with the full SAM index (thick green line in Figure~\ref{fig:sam-eof-vertical}), although not statistically significant at every level.

\textcolor{red}{(Esta explicación habría que hacerla más detallada. Explicar más sobre lo que se ve en cada panel y después ir interpretando.  )}
The split between A-SAM and S-SAM gives more insight into the nature of the relationship. The relationship between the SAM and ¿the imaginary cEOF1? (Figure~\ref{fig:sam-eof-vertical}.b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.

The Imaginary cEOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} -96\% of shared variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).

\begin{figure}
\centering
\includegraphics{../figures/eof-filtered-1.pdf}
\caption{\label{fig:eof-filtered}Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º. Arbitrary units.}
\end{figure}

@@

As a further illustration \textcolor{red}{(¿De qué?)}, Figure~\ref{fig:eof-filtered} shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed. The resulting pattern is an nearly zonally symmetrical annular mode. This mode is highly correlated with SAM (0.75 (CI: 0.58 -- 0.86)), but only with the symmetrical part (0.88 (CI: 0.79 -- 0.94)). Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.

\begin{figure}
\centering
\includegraphics{../figures/fake-eof-1.pdf}
\caption{\label{fig:fake-eof}EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the cEOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.}
\end{figure}

This suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA. Figure~\ref{fig:fake-eof} serves as an illustration of this kind of statistical mixing of independent modes. If shows the leading (and only) EOFs of a synthetic dataset if 41 years created by adding a perfectly zonally symmetrical SAM-like pattern with random amplitudes taken from the real SAM, and geopotential height anomalies predicted by EFO2. Even though by construction this dataset has a perfectly symmetric SAM, the EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.

@@

Since the Real cEOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location. However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics. The correlation between the Real cEOF2 and ENSO is also not significant () \textcolor{red}{(No aparece el valor en el PDF)}. This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.

@@

Furthemore, Figure~\ref{fig:enso-phase} suggest that strong cEOF2 years tend to coincide with strong ENSO years. The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.44 (CI: 0.16 -- 0.66). This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest cEOF2 magnitude. If those years are removed, the correlation becomes non-significant (0.044 (CI: -0.28 -- 0.36)). Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.2). Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.

>>>>

\begin{figure}
\centering
\includegraphics{../figures/eof-naive-1.pdf}
\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units).}
\end{figure}

@@

This is similar to the method used by \citet{irving2016}. From their Fourier decomposition of reprojected meridional wind fields, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peacks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).

@@

\begin{figure}
\centering
\includegraphics{../figures/sam-eof-vertical-1.pdf}
\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}. Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
\end{figure}

To explore the relationship between SAM and the modes described from the cEOF, we compute. the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.~\ref{fig:sam-eof-vertical}). The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere. It is not correlated with the Real cEOF2 at any level.

The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa. In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM. These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh}.

@@

In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered. The Imaginary cEOF2 and the A-SAM share up to 92\% of their variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2). Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these are not different features that share a causal link, but that they might be different ways of characterising the same phenomenon. And, in fact, suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA.

\begin{figure}
\centering
\includegraphics{../figures/fake-eof-1.pdf}
\caption{\label{fig:fake-eof}Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa. Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis. Units are arbitrary. Negative values in blue and positive values in red.}
\end{figure}

To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of 41 fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value. To this synthetic dataset we then apply traditional EOF. By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA of varying amplitude and phase (Figure~\ref{fig:fake-eof} row a). The former one is ar standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly. Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.

The three EOFs of the synthetic dataset are shown in Figure~\ref{fig:fake-eof} row b. The synthetic SAM can be seen in the first EOF while the PSA is explained by the second and third. However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not. The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern. Of note, the three EOFs are well separated according to North's rule of thumb \citep{north1982}.

@@

Since the Real cEOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location. However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics. The correlation between the Real cEOF2 and ENSO is also not significant (-0.0017 (CI: -0.31 -- 0.31)) \textcolor{red}{(No aparece el valor en el PDF)}. This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.

@@

Furthermore, Figure~\ref{fig:enso-phase} suggest that strong cEOF2 years tend to coincide with strong ENSO years. The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.44 (CI: 0.16 -- 0.66). This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest cEOF2 magnitude. If those years are removed, the correlation becomes non-significant (0.044 (CI: -0.28 -- 0.36)). Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.2). Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
